---
title: "Impact of Weather Events on Public Health and Economics in US"
output: 
  html_document:
    keep_md: true
---

## Synopsis
Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern. This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm dataset. The results show that `tornado` is the most harmful weather event to public health. As for the economic loss, `flood` causes the largest property loss and `drought`  makes the largest crop damage.

## Data Processing
The events in the dataset start in the year 1950 and end in November 2011. In the earlier years of the dataset there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete. The data used in this report are available [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) [47Mb].  


After loading data from bzip2 file, one can find that the types of events are stored in **`EVTYPE`** variable,the fatalities are stored in **`FATALITIES`**, the injuries are stored in **`INJURIES`**, the property damage is stored in **`PROPDMG`** and **`PROPDMGEXP`**, and the crop damage is stored in **`CROPDMG `** and **`CROPDMGEXP`**.

```{r, cache=TRUE, results='hide'}
Sys.setlocale("LC_ALL", "C")
data = read.csv(bzfile("repdata-data-StormData.csv.bz2"),stringsAsFactors=FALSE)
```

Before analysis, we convert the items in **`EVTYPE`** to lowercase and replace all punctuation and characters with a space.

```{r}
data2 = data
# before processing
length(unique(data2$EVTYPE)) 
```



```{r}
data2$EVTYPE = tolower(data2$EVTYPE)
data2$EVTYPE = gsub("[[:blank:][:punct:]+]", " ", data2$EVTYPE)
# after processing
length(unique(data2$EVTYPE)) 
```

One can see that after the pre-processing, the number of unique event types reduce significantly(985 V.S. 874).

 
In addtion to pre-processing `EVTYPE`, we also need to deal with the exponents of `PROPDMGEXP` and `CROPDMGEXP`. The exponents of `PROPDMGEXP` and `CROPDMGEXP` are shown as follow.

```{r}
# PROPDMGEXP
unique(data2$PROPDMGEXP)
# CROPDMGEXP
unique(data2$CROPDMGEXP)
```

To deal with this exponent issue, we build a `Converter` function to transfer the exponent into a number.

```{r}
Converter <- function(e) {
  # h -> hundred, k -> thousand, m -> million, b -> billion
  if (e %in% c('h', 'H'))
    return(2)
  else if (e %in% c('k', 'K'))
    return(3)
  else if (e %in% c('m', 'M'))
    return(6)
  else if (e %in% c('b', 'B'))
    return(9)
  else if (!is.na(as.numeric(e))) # if a digit
    return(as.numeric(e))
  else 
    return(0)
}
```

## Results

### How Weather Events affect Public Health

Here is the top 20 events of the highest fatalities. 
```{r, cache=TRUE}
# weather events causes death
FATALITIES = head(sort(tapply(data2$FATALITIES,data2$EVTYPE, sum),decreasing=TRUE), n=20)
FATALITIES = data.frame("Event"=names(FATALITIES),"Fatalities"=unname(FATALITIES))
FATALITIES
```

```{r, cache=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(scales)

ggplot(data=FATALITIES,
             aes(x=reorder(Event,Fatalities), y=Fatalities, fill=Fatalities)) +
  geom_bar(stat="identity") +
  coord_flip() +
  ylab("Total number of fatalities") +
  xlab("Event type") +
  theme(legend.position="none") + 
  ggtitle("Top deadly weather events in the US (1950-2011)") +
  scale_y_log10(breaks = 
                  trans_breaks("log10", function(x) 10^x), labels = trans_format("log10", math_format(10^.x)))

```

Here is the top 20 events of the highest injuries.
```{r, cache=TRUE}
# weather events causes injuries
INJURIES = head(sort(tapply(data2$INJURIES,data2$EVTYPE, sum),decreasing = TRUE), n=20)
INJURIES = data.frame("Event"=names(INJURIES),"Injuries"=unname(INJURIES))
INJURIES
```


**It is obvious that the tornado is the most harmful weather event to population health, which causes more than 5,000 death and  more than 10,000 injuries in the last 60 years.** 


### How Weather Events affect Economics

First, we combine the damage loss and the damage loss exponent together, i.e. $$PROPDMG^{PROPDMGEXP}\ \ and\ \    CROPDMG^{CROPDMGEXP},$$ to get the total loss in dollars. 

```{r, cache=TRUE, warning=FALSE, message=FALSE}
PROPDMGEXP <- sapply(data2$PROPDMGEXP, Converter)
data2$PROPDMG <- data2$PROPDMG*(10^unname(PROPDMGEXP))
CROPDMGEXP <- sapply(data2$CROPDMGEXP, Converter)
data2$CROPDMG <- data2$CROPDMG * (10^unname(CROPDMGEXP))
```


Here is the top 20 events of the highest property damage. 
```{r, cache=TRUE}
# weather events causes death
PROPDMG = head(sort(tapply(data2$PROPDMG,data2$EVTYPE, sum),decreasing=TRUE), n=20)
PROPDMG = data.frame("Event"=names(PROPDMG),"PROPDMG"=unname(PROPDMG))
PROPDMG
```

```{r, cache=TRUE, warning=FALSE, message=FALSE}
ggplot(data=PROPDMG,
             aes(x=reorder(Event,PROPDMG), y=PROPDMG, fill=PROPDMG)) +
  geom_bar(stat="identity") +
  coord_flip() +
  ylab("Property damage in dollars") +
  xlab("Event type") +
  theme(legend.position="none") + 
  ggtitle("Weather costs to the US economy (1950-2011)") +
  scale_y_log10(breaks = 
                  trans_breaks("log10", function(x) 10^x), labels = trans_format("log10", math_format(10^.x)))

```

Here is the top 20 events of the highest crop damage.
```{r, cache=TRUE}
# weather events causes injuries
CROPDMG = head(sort(tapply(data2$CROPDMG,data2$EVTYPE, sum),decreasing = TRUE), n=20)
CROPDMG = data.frame("Event"=names(CROPDMG),"CROPDMG"=unname(CROPDMG))
CROPDMG
```

```{r, cache=TRUE, warning=FALSE, message=FALSE}
ggplot(data=CROPDMG,
             aes(x=reorder(Event,CROPDMG), y=CROPDMG, fill=CROPDMG)) +
  geom_bar(stat="identity") +
  coord_flip() +
  ylab("Crop damage in dollars") +
  xlab("Event type") +
  theme(legend.position="none") + 
  ggtitle("Weather costs to the US economy (1950-2011)") +
  scale_y_log10(breaks = 
                  trans_breaks("log10", function(x) 10^x), labels = trans_format("log10", math_format(10^.x)))
```

**The figures show that the flood causes the largest property loss while the drought makes the largest crop damage.** 